\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{placeins}
\usetikzlibrary{shapes.geometric, arrows, positioning}
\setlength{\emergencystretch}{3em}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
    pdftitle={Literature Review: Toward Signal-Preserving TinyML for Gravitational Wave Triggers},
    pdfpagemode=FullScreen,
}

\title{\textbf{Literature Review: Toward Signal-Preserving TinyML for Gravitational Wave Triggers} \\
\large Resolving the Tension Between Astrophysical Fidelity and Sub-Watt Model Compression}
\author{\textbf{Razeen Wasif} \\ Uni ID: u7283652 \\ Course Code: COMP4450}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Detection of compact-binary coalescences within the millisecond budget of multi-messenger astronomy is now constrained by hardware as much as by algorithms. Eighteen papers from gravitational-wave (GW) machine learning, embedded inference, and quantisation theory are synthesised to ask whether the trigger stage of a GW pipeline can be pushed onto sub-watt microcontrollers without losing astrophysical sensitivity. The literature separates into two weakly connected camps. One has shown that deep networks can match matched filtering on high-power GPUs. The other has shown that aggressive compression is feasible, but evaluates it on images and audio rather than on chirp waveforms. The hinge between the two is INT8 quantisation, which acts on a phase-coherent waveform as a structured perturbation rather than the well-behaved rounding error assumed by computer-vision benchmarks. Existing quantisation-aware training and distillation work does not address this, leaving a clear opening for physics-informed, signal-preserving compression.
\end{abstract}

\noindent\textbf{Keywords:} TinyML; quantisation-aware training; gravitational-wave detection; signal-to-noise ratio; embedded machine learning; low-latency triggers; multi-messenger astronomy.

\smallskip
\noindent\textbf{ACM CCS:} Computing methodologies $\rightarrow$ Machine learning algorithms; Computer systems organization $\rightarrow$ Embedded systems; Applied computing $\rightarrow$ Astronomy.

\section{Introduction}

\subsection{Summary of the Research Proposal}
This review provides the literature foundation for the COMP4450 Research Proposal \textit{``Quantisation-Induced SNR Degradation: Toward Signal-Preserving TinyML for Gravitational Wave Triggers''} (Wasif 2026). The proposal sits at the intersection of three computer-science sub-fields. \textit{Intelligent Systems} is the primary home; \textit{Embedded Computer Systems} and \textit{Computational Astrophysics} are secondary contexts. Its central claim is that next-generation gravitational-wave detectors face a data-deluge problem GPU-bound deep-learning pipelines cannot solve. Missions such as LISA and CubeSat-scale autonomous arrays simply cannot downlink the raw data, so the trigger stage has to move onto microcontrollers running in 128~KB of SRAM at sub-watt power. That move forces aggressive INT8 quantisation, and the proposal argues this will destroy the phase coherence of the chirp waveform. The research question follows directly: to what extent does non-linear bit-width reduction interfere with the phase coherence on which reliable GW triggering depends?

\subsection{Aim and research questions}
The aim of this review is to survey the literature that approaches the gap from two directions: GW machine learning on one side, and TinyML and quantisation theory on the other. The case made here is that the intersection remains unresolved. Two questions structure the survey. \textit{RQ1} asks how prior work has traded model size, latency, and astrophysical sensitivity. \textit{RQ2} asks which quantisation-aware techniques, if any, have been validated against astrophysical metrics such as false alarm rate and distance reach, rather than against the top-1-accuracy metrics inherited from computer vision. The technical context matters here. Detecting strains of $h \approx 10^{-21}$ by matched filtering is statistically optimal but computationally heavy (Abbott et al. 2020), and template-bank growth is now the limiting factor in multi-messenger follow-up (Cuoco et al. 2024). Deep learning matches matched-filter sensitivity in published results (George and Huerta 2018; Gabbard et al. 2018), but those results all assume FP32 arithmetic. What happens when that arithmetic is compressed to INT8 is the question this review traces.

\section{Materials and Methods}

A targeted, reproducible review was conducted between March and April 2026. It is lighter than a formal systematic review, but disciplined enough that someone else could re-run the search.

\subsection{Search strategy}
Four databases were used: arXiv (gr-qc, cs.LG, eess.SP), IEEE Xplore, the ACM Digital Library, and Google Scholar. Query strings paired the GW side (\textit{``gravitational wave'' OR ``binary black hole'' OR ``binary neutron star''}) with the ML and hardware side (\textit{``deep learning'' OR ``convolutional'' OR ``quantisation'' OR ``FPGA'' OR ``microcontroller'' OR ``TinyML''}). The Gravitational Wave Open Science Center (GWOSC) was checked separately for datasets and observational papers. Three seed papers were used for forward and backward citation chasing: George and Huerta (2018), Gebhard et al. (2019), and Gholami et al. (2021).

\subsection{Inclusion and exclusion criteria}
Papers were included if they did one of three things. First, propose or evaluate a machine-learning model for GW detection, denoising, or trigger generation. Second, supply foundational technique work on quantisation, distillation, or embedded inference frameworks the GW side draws on. Third, report quantitative results on either sensitivity (SNR, FAR, detection probability) or efficiency (latency, memory, parameter count). Papers were excluded if they focused only on offline Bayesian parameter estimation, on cosmology without a trigger stage, or used neural networks merely for post-hoc visualisation. Murphy et al.\ (2021) on the EIRSAT-1 CubeSat gamma-ray detector is included on the deployment side as the only autonomous, downlink-constrained instrument paper in the set with directly comparable on-board triggering constraints.

\subsection{Inclusion outcome}
Eighty-six records were identified. After deduplication and title/abstract screening 32 remained for full-text assessment, and 18 were finally included for in-depth review (Figure \ref{fig:prisma}). The included set divides into four thematic groups: seven papers showing ML matching matched filtering, four on quantisation theory and embedded frameworks, three on FPGA-class hardware deployment in physics, and four critical or foundational papers on noise, glitches, and benchmarks.

\subsection{Validity}
Three measures support reproducibility. First, every included paper is cited with a DOI or arXiv ID, so the search can be replicated in five years. Second, a ``clique map'' was maintained that assigns each paper to one or more of three thematic axes (sensitivity, efficiency, robustness). The map makes visible where the literature is thick and where it is thin. Third, claims were cross-checked between independent works. The adversarial-noise critique of George and Huerta (2018) by Gebhard et al. (2019) is one example, and Banbury et al.'s (2021) microcontroller benchmarks, when read against the SRAM budgets assumed by David et al. (2021), are another.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.6cm,
  box/.style={draw, rectangle, minimum width=6.5cm, minimum height=1cm, align=center, fill=gray!5, font=\footnotesize},
  arrow/.style={thick,->,>=stealth}
]
  \node (ident) [box] {\textbf{Identification} \\ Records identified: arXiv (54) $\cdot$ IEEE (12) $\cdot$ ACM (4) $\cdot$ Scholar (16) \\ Total = 86};
  \node (screen) [box, below=of ident] {\textbf{Screening} \\ After deduplication: 71 \\ Title/abstract screened: 71 \\ Excluded (off-topic): 39};
  \node (elig) [box, below=of screen] {\textbf{Eligibility} \\ Full text assessed: 32 \\ Excluded (no quant. results, offline PE, no GW/compression): 14};
  \node (incl) [box, below=of elig] {\textbf{Included} \\ Papers included for in-depth synthesis: 18 \\ (sensitivity 7 $\cdot$ efficiency 4 $\cdot$ hardware 3 $\cdot$ noise/benchmarks 4)};

  \draw [arrow] (ident) -- (screen);
  \draw [arrow] (screen) -- (elig);
  \draw [arrow] (elig) -- (incl);
\end{tikzpicture}
\caption{PRISMA-style flow of identification, screening, eligibility, and inclusion.}
\label{fig:prisma}
\end{figure}

\section{Results}

\noindent\textbf{Key message.} The literature splits into two cliques that rarely cite each other. A sensitivity-first clique (George and Huerta 2018; Gabbard et al. 2018; Krastev 2020; Schäfer et al. 2023; Dax et al. 2021; Chaturvedi et al. 2022) shows that deep learning can match matched filtering, but only on GPU. A deployment-first clique (David et al. 2021; Banbury et al. 2021; Gholami et al. 2021; Jacob et al. 2018; Hinton et al. 2015) shows that aggressive compression is feasible, but evaluates models on images, audio, or generic benchmarks rather than on astrophysical strain. Sitting between the two are critical papers (Gebhard et al. 2019; Abbott et al. 2016; Abbott et al. 2020) that warn the resulting models are fragile in ways simple accuracy metrics cannot detect. Mapped onto contribution axes the 18 papers split roughly seven on sensitivity, six on efficiency, and three on robustness; only Qi et al.\ (2024) and Chaturvedi et al.\ (2022) contribute primarily on more than one, and no paper contributes primarily on all three. Figure \ref{fig:bitwidth} projects the same literature onto the axes the proposal's hypothesis actually lives on: weight precision and sensitivity preservation.

\subsection{The sensitivity-first clique}
George and Huerta (2018) and Gabbard et al.\ (2018) form the foundation of ML-based GW detection. Both show that a relatively shallow one-dimensional CNN can recover binary black hole (BBH) signals from Advanced LIGO strain at sensitivity comparable to matched filtering. They are best read as a clique: they converge on the same architectural choice (deep 1-D convolutions over whitened strain) and report the same null result (no measurable loss of true-positive rate at high SNR), so their joint claim is stronger than either paper alone.

Krastev (2020) extended the design to binary neutron star (BNS) signals, and in doing so exposed a weakness the BBH papers had concealed. BNS waveforms are an order of magnitude longer in the detector band, so the receptive field of any TinyML candidate model has to grow proportionally. That requirement runs directly into the SRAM budget assumed by David et al. (2021).

Schäfer et al. (2023) formalised the comparison through the first ML mock-data challenge. The headline result was that ML pipelines were competitive with PyCBC at high false-alarm-rate thresholds, but lagged at the low-FAR regime that astrophysical confirmation actually requires. Dax et al. (2021) and Chaturvedi et al. (2022) close out the clique on the inference side. The former demonstrates real-time \textit{posterior} estimation in seconds. The latter applies NVIDIA TensorRT mixed-precision (FP16/INT8) layer fusion and reports processing approximately one month of LIGO data in about 50 seconds, a roughly $3\times$ speedup, with negligible sensitivity loss for known BBH events. The limitation is that this result depends on A100-class datacentre GPUs and TensorRT calibration, not on memory-static microcontroller inference. The collective blind spot of the clique is therefore hardware scale.

\subsection{The deployment-first clique}
Where the sensitivity clique assumes GPUs, the deployment clique assumes microcontrollers. David et al. (2021) define the de facto runtime for TinyML in the form of TensorFlow Lite Micro, and lay out the engineering primitives that govern what can and cannot run in approximately 256~KB of SRAM: a single pre-allocated tensor arena, op-by-op execution, no dynamic allocation after initialisation, and vendor-specific kernels such as CMSIS-NN. Their evaluation reports interpreter overhead below 1\%. Banbury et al. (2021) supply the calibrating benchmark suite (MLPerf Tiny). Both target keyword spotting, image classification, and anomaly detection, not astrophysical time series.

The theoretical machinery sits with Gholami et al. (2021), whose taxonomy of post-training quantisation (PTQ) and quantisation-aware training (QAT) is the canonical reference, and with Jacob et al. (2018), whose integer-arithmetic-only inference paper supplies the mathematical foundation, including the straight-through estimator for backpropagating through a discontinuous rounding operator. Knowledge distillation (Hinton et al. 2015) is the complementary tool: a high-precision teacher transfers softened class probabilities to a quantised student, recovering accuracy PTQ discards.

The clique is internally coherent and well-cited. But its evaluation regime is uniformly computer-vision-centric: top-1 ImageNet accuracy, word-error rate, mean average precision. None of these metrics quantifies the loss of phase coherence that matters for a chirp signal. That gap is the central observation of this review.

\subsection{Hardware-software co-design on FPGA-class targets}
A small group of papers sits between the two cliques and provides the closest evidence for the proposal's hardware claim. Qi et al. (2024) implemented a CNN for BNS early-warning on an FPGA, achieved sub-millisecond inference, and showed empirically that INT8 quantisation was tolerable at stronger signal regimes but degraded sensitivity for the long-duration BNS signals that motivated the work. Within the included set, this is the clearest empirical evidence for the SNR--precision trade-off the proposal hypothesised. Chaturvedi et al. (2022) span both cliques. Their TensorRT result is on GPU, but it operationalises the same mixed-precision idea microcontroller deployments will eventually need: higher precision in sensitive layers, INT8 elsewhere. The two papers are the closest existing analogues to ``signal-preserving TinyML''. Neither, however, tests a sub-watt microcontroller target, reports peak SRAM under a TinyML memory arena, uses astrophysical distance reach as a metric, or generalises the underlying framework.

\subsection{Noise, glitches, and the brittleness of ML triggers}
Gebhard et al. (2019) is the critical pivot of the review. It re-examines the George and Huerta (2018) and Gabbard et al. (2018) results, and shows the high reported accuracies disguise something the authors did not test for: adversarial brittleness. Small, structured perturbations to the input strain flip the network output with high confidence. More importantly for the present review, the network confidently classifies inputs that bear no visible resemblance to a chirp. Gebhard et al. conclude that CNNs should be used as trigger generators only, not as confirmation tools.

Abbott et al. (2016) catalogue the morphology of ``blip'' transients in real LIGO data. These are exactly the kind of high-amplitude, short-duration response a quantised CNN is most likely to misclassify. Abbott et al. (2020) document the data-quality vetoes the LVK collaboration uses to mitigate them offline. The implication is uncomfortable for TinyML. If FP32 networks are already adversarially fragile, then INT8 quantisation, which is itself a structured, deterministic perturbation of the network's response surface, is more likely to amplify that fragility than to be neutral to it. None of the deployment-first papers acknowledge this adversarial-noise literature. That is a serious omission.

\subsection{Toward signal-preserving TinyML}
Cuoco et al.\ (2021) provide the earlier of two field-shaping reviews of ML in GW science; Cuoco et al.\ (2024) extend that coverage to current detectors. Neither devotes more than a paragraph to sub-watt deployment. The methodological building blocks for closing the gap already exist. Hinton et al.'s (2015) distillation framework supplies the training objective. Jacob et al.'s (2018) integer-only inference and Gholami et al.'s (2021) QAT taxonomy supply the quantisation machinery. What the literature does not contain is a composition of those blocks under astrophysical, rather than computer-vision, loss functions. Murphy et al.'s (2021) work on the EIRSAT-1 CubeSat gamma-ray detector illustrates the deployment niche the proposal is aimed at: a compact, autonomous, downlink-constrained instrument where on-board triggering is operationally required.

\subsection{The wrong benchmark}
The sensitivity-first clique evaluates ML detectors by parity with matched filtering at fixed FAR. That benchmark is appropriate for a confirmation pipeline, but a trigger is not a confirmation pipeline. In a production GW analysis the trigger only needs to flag candidate windows cheaply enough that a second, higher-precision stage can re-do the heavy work. This trigger-then-confirm hierarchy is not hypothetical: the GstLAL low-latency pipeline (Messick et al.\ 2017) is explicitly designed as a rapid trigger that feeds deeper offline analyses, so the reframing argued here aligns with how the LVK collaboration already operates rather than inventing a new structure. Matched-filter parity at the trigger stage is therefore over-specification, and the FP32 assumption that lets the clique meet that bar is exactly what blocks the move to TinyML. Reading Qi et al.\ (2024) through this lens changes the interpretation of their INT8 BNS result. The reported sensitivity loss is a failure only against the matched-filter-parity benchmark. Against a hierarchical-trigger benchmark, where the relevant metric is candidate recall conditional on a tractable downstream load, the same result is a partial success, provided the trigger preserves enough phase information to feed the second stage. This is the operational meaning of ``signal-preserving'' that the literature has not articulated, and it is the reason the empty region in Figure \ref{fig:bitwidth} is empty: the clique that should be filling it is optimising for the wrong target.

\begin{figure}[!htbp]
\centering
\begin{tikzpicture}[font=\footnotesize]

  % --- axes ---
  \draw[->, thick] (0,0) -- (10.5,0);
  \draw[->, thick] (0,0) -- (0,6.4);

  % x-axis ticks (log-spaced bitwidths)
  \foreach \x/\bits in {1/4, 3/8, 6/16, 9/32} {
    \draw[thick] (\x,0) -- (\x,-0.15);
    \node[below] at (\x,-0.2) {\bits};
  }
  \node[below, align=center] at (5,-0.8) {weight precision (bits)};

  % y-axis ticks (sensitivity preserved vs FP32 baseline)
  \foreach \y/\pct in {1.0/50, 2.5/70, 4.0/85, 5.5/100} {
    \draw[thick] (-0.15,\y) -- (0,\y);
    \node[left] at (-0.2,\y) {\pct\%};
  }
  \node[rotate=90, anchor=south] at (-1.4,3.2) {sensitivity preserved vs FP32 baseline};

  % FP32 baseline line
  \draw[dashed, gray!60] (0,5.5) -- (10,5.5);
  \node[gray, right, font=\scriptsize] at (9.5,5.7) {FP32};

  % --- target / empty region ---
  \fill[green!15] (0.3,4.4) rectangle (3.6,5.6);
  \draw[green!50!black, thick, dashed] (0.3,4.4) rectangle (3.6,5.6);
  \node[green!40!black, font=\scriptsize\bfseries, align=center] at (1.95,6.05) {empty region};
  \node[green!40!black, font=\scriptsize, align=center] at (1.95,4.15) {(sub-watt INT8 BNS)};

  % --- BBH curve (solid blue) ---
  \draw[blue, thick] plot[smooth, tension=0.5] coordinates {(9,5.5) (6,5.45) (3,5.30) (1,3.6)};
  \node[blue, font=\scriptsize] at (7.4,5.25) {BBH (reported)};

  % --- BNS curve (dashed red) ---
  \draw[red, thick, dashed] plot[smooth, tension=0.5] coordinates {(9,5.5) (6,5.05) (3,3.0) (1,1.0)};
  \node[red, font=\scriptsize] at (7.4,4.7) {BNS (conjectured)};

  % --- paper anchor points ---
  % BBH at FP32: George & Huerta 2018, Gabbard 2018
  \fill[blue] (9,5.5) circle (3pt);
  \node[blue, above right, font=\scriptsize] at (9.05,5.55) {G\&H, Gabbard};

  % BNS at FP32: Krastev
  \fill[red] (9,5.42) circle (3pt);
  \node[red, below right, font=\scriptsize] at (9.05,5.30) {Krastev};

  % Chaturvedi 2022: mixed FP16/INT8 BBH
  \fill[blue] (4,5.36) circle (3pt);
  \node[blue, above, font=\scriptsize] at (4,5.55) {Chaturvedi};

  % Qi 2024: INT8 BNS FPGA, reported degradation
  \fill[red] (3,3.0) circle (3pt);
  \node[red, right, font=\scriptsize] at (3.15,2.95) {Qi 2024};

  % Jacob 2018: INT8 ImageNet (cross-domain reference)
  \fill[gray!70] (3,5.30) circle (3pt);
  \node[gray, below, font=\scriptsize] at (3,5.0) {Jacob (CV ref.)};

\end{tikzpicture}
\caption{Synthesis of reported and conjectured sensitivity preservation as a function of weight precision, drawn from the 18 included papers. BBH classification (blue, solid) holds close to the FP32 baseline through INT8 in both the GW literature (Chaturvedi et al.\ 2022, TensorRT mixed FP16/INT8) and the wider quantisation literature (Jacob et al.\ 2018 on ImageNet, plotted in grey as a cross-domain reference). BNS sensitivity (red, dashed) degrades more sharply: Qi et al.\ (2024) is the single empirical INT8 anchor in the GW literature, and the extrapolation toward 4-bit follows the proposal's phase-coherence hypothesis. The shaded green region (INT8 or below, BNS-class sensitivity preserved, on a sub-watt microcontroller budget) is the target operating zone and contains no published work in the included set. The figure is a synthesis sketch, not an original experimental result.}
\label{fig:bitwidth}
\end{figure}

\noindent The empty green region in Figure \ref{fig:bitwidth}, not the absence of any single paper, is the research gap PC-KD (\S4.3) is designed to fill.

\FloatBarrier

\section{Discussion}

\noindent\textbf{Key message.} The literature contains every ingredient required for signal-preserving TinyML, but no published work composes those ingredients under astrophysical loss functions. The single FPGA-class deployment in the included set (Qi et al. 2024) already shows that naive INT8 erodes BNS sensitivity. The gap is therefore one of evaluation discipline as much as of technique.

\subsection{Principal results}
Four findings emerge. First, deep learning has reached parity with matched filtering for high-SNR BBH signals (George and Huerta 2018; Gabbard et al.\ 2018; Schäfer et al.\ 2023), but the parity is fragile at low FAR and on long-duration BNS waveforms. Second, the embedded-ML literature measures the cost of quantisation in classification accuracy (Jacob et al.\ 2018; Gholami et al.\ 2021; Banbury et al.\ 2021), a metric that is structurally insensitive to phase-coherent SNR loss. Third, the only FPGA-class GW deployment in the included set (Qi et al.\ 2024) demonstrates that naive quantisation degrades sensitivity in exactly the regime that edge triggers were meant to enable. Fourth, the matched-filter-parity benchmark on which the sensitivity-first clique is judged is itself the wrong target for an edge trigger, which operates as a cheap first stage in a hierarchical pipeline; under that reframing Qi et al.\ (2024) is a partial success rather than a failure.

\subsection{Limitations}
Three limitations should be flagged. The 18-paper cap forced the exclusion of several adjacent works, in particular recent mixed-precision FPGA inference outside astronomy and the emerging transformer-based GW pipelines. These are present in the discipline but not in the in-depth synthesis here. Some of the most recent work cited is at arXiv-preprint stage, which means peer-review status varies. The synthesis in Figure \ref{fig:bitwidth} is a coarse projection of a multi-dimensional problem onto two axes: it suppresses activation precision, per-layer mixed-precision schemes, calibration choices, and the heterogeneity inside the BBH and BNS categories. The BNS curve in particular is hypothesised from a single empirical anchor (Qi et al.\ 2024) and will need to be re-drawn as more INT8 BNS deployments are published.

\subsection{Proposed future work: Phase-Coherence-Weighted Knowledge Distillation}
The empty region of Figure \ref{fig:bitwidth} and the reframing of \S3.6 motivate a specific named contribution that this review identifies as the methodological piece the literature is missing. \textit{Phase-Coherence-Weighted Knowledge Distillation} (PC-KD) is a distillation objective in which the student loss is weighted sample-by-sample by the time-frequency phase coherence of the input window against a chirp template. Quantisation noise injected in low-coherence regions of the signal is penalised softly. Quantisation noise injected at the merger ringdown, where phase information dominates SNR, is penalised hard. Concretely,
\begin{equation}
\mathcal{L}_{\text{PC-KD}}(x) \;=\; \alpha \, \mathrm{KL}\big( p_T(x) \,\|\, p_S(x) \big) \;+\; \beta \, w(x) \, \big\| f_T(x) - f_S(x) \big\|^2,
\label{eq:pckd}
\end{equation}
where $p_T, p_S$ are teacher and student class probabilities, $f_T, f_S$ are their penultimate-layer feature maps (after Hinton et al.\ 2015), and $w(x)$ is an instantaneous phase-coherence weight derived from a matched-filter pre-pass against a coarse template bank. The four research directions implied by the gap analysis are best understood as components of building PC-KD: an astrophysical loss in place of cross-entropy is the first term of Equation \ref{eq:pckd}; a microcontroller-class mixed-precision deployment after Chaturvedi et al.\ (2022) is the target the student is compressed for; the adversarial-robustness diagnostics of Gebhard et al.\ (2019) are the stress test the student must pass under real LIGO glitches (Abbott et al.\ 2016; Abbott et al.\ 2020); and an MLPerf-Tiny-style benchmark (Banbury et al.\ 2021) parameterised by astrophysical reach is the yardstick by which PC-KD is scored against the naive INT8 baseline of Qi et al.\ (2024).

\subsection{Comparison with prior work}
Cuoco et al.\ (2021; 2024) are the most comprehensive prior reviews of ML in GW science, but both treat hardware deployment as downstream engineering and devote little space to bit-width effects. On the TinyML side, Warden and Situnayake (2019), Lin et al.\ (2020) on MCUNet, and Sze et al.\ (2017) on efficient hardware inference survey the embedded landscape with equal authority and no astrophysical content. This review differs in two ways. It occupies the intersection of the two literatures and treats the intersection itself as the problem, with Gebhard et al.\ (2019) and Qi et al.\ (2024) carrying the load. And it goes beyond surveying: \S3.6 reframes the matched-filter-parity benchmark of the sensitivity-first clique, and \S4.3 proposes PC-KD as a named methodological contribution rather than a generic ``future work'' bullet.

\subsection{Conclusions}
Putting TinyML into gravitational-wave detection is a prerequisite for the next generation of multi-messenger observatories. It cannot be done by stacking off-the-shelf compression on top of off-the-shelf ML pipelines, and it cannot be done while the sensitivity-first clique optimises for matched-filter parity instead of trigger-stage recall. Every ingredient already exists: architectures that match matched filtering, runtimes that fit in 256~KB, quantisation procedures proven on millions of devices, distillation that preserves teacher knowledge. What does not yet exist is a composition of those ingredients under astrophysical metrics. Phase-Coherence-Weighted Knowledge Distillation, as proposed in \S4.3 and adopted by Wasif (2026), is the missing piece this review identifies.

\section{Self-Reflection}
Generative AI was used in three separate roles during the preparation of this review: Google Gemini CLI throughout, and a single cross-checking pass with Anthropic Claude on the near-final draft. The first role was drafting. Zero-shot prompts produced rough section skeletons that were then heavily rewritten by hand. The second role was reference discovery, and this is where the work demanded the most discipline. An initial prompt (``list TinyML papers for gravitational wave detection'') returned a confident, plausible-looking list in which several of the DOIs turned out to be fabricated. Refining the prompt to ``list peer-reviewed FPGA implementations of CNNs for binary-neutron-star detection with arXiv IDs verifiable on arXiv.org'' returned a much shorter, more honest list, of which Qi et al.\ (2024) was the only verifiable hit. The third role was copy-editing on near-final paragraphs. The clear advantage of GenAI in this work was the speed of cross-walking between gravitational-wave physics and embedded-systems literature, two sub-disciplines that almost never cite each other. The clear disadvantage was the model's willingness to fabricate references, which on first contact would have produced a serious academic-integrity failure. The mitigation, and the lesson recorded for future work, was that every model-suggested reference was independently verified, and no claim was kept in the final draft without an author-checked source. One subtler failure mode became visible only late: when asked to summarise a paper I had not yet read, the model tended to over-state the strength of the paper's conclusions in ways quiet enough to survive casual review. The corrective for future literature work is to draft summaries from the PDFs first, and to use the model only for comparison or copy-editing afterwards.

\section*{List of References}

\begin{enumerate}[label=\arabic*., leftmargin=2em]
\item Abbott, BP et al.\ 2016, `Characterization of transient noise in Advanced LIGO relevant to gravitational-wave signal GW150914', \textit{Classical and Quantum Gravity}, vol.\ 33, no.\ 13, 134001, doi:10.1088/0264-9381/33/13/134001. \url{https://doi.org/10.1088/0264-9381/33/13/134001}.*
\item Abbott, BP et al.\ 2020, `A guide to LIGO--Virgo detector noise and extraction of transient gravitational-wave signals', \textit{Classical and Quantum Gravity}, vol.\ 37, no.\ 5, 055002, doi:10.1088/1361-6382/ab685e. \url{https://doi.org/10.1088/1361-6382/ab685e}.*
\item Banbury, CR, Reddi, VJ, Lam, P, Fu, W, Fazel, A, Holleman, J, Huang, X, Hurtado, R, Kanter, D, Lokhmotov, A, Patterson, D, Pau, D, Seo, J, Sieracki, J, Thakker, U, Verhelst, M \& Yadav, P 2021, `MLPerf Tiny benchmark', in \textit{Proceedings of the 35th Conference on Neural Information Processing Systems (NeurIPS) Datasets and Benchmarks Track}, arXiv:2106.07597. \url{https://arxiv.org/abs/2106.07597}.*
\item Chaturvedi, P, Khan, A, Tian, M, Huerta, EA \& Zheng, H 2022, `Inference-optimized AI and high performance computing for gravitational wave detection at scale', \textit{Frontiers in Artificial Intelligence}, vol.\ 5, 828672, doi:10.3389/frai.2022.828672. \url{https://doi.org/10.3389/frai.2022.828672}.*
\item Cuoco, E, Powell, J, Cavagli\`{a}, M, Ackley, K, Bejger, M, Chatterjee, C, Coughlin, M, Coughlin, S, Easter, P, Essick, R, Gabbard, H, Gebhard, T, Ghosh, S, Haegel, L, Iess, A, Keitel, D, M\`{a}rka, Z, M\`{a}rka, S, Morawski, F, Nguyen, T, Ormiston, R, P\"{u}rrer, M, Razzano, M, Salemi, F, Sasaoka, S \& Williamson, A 2021, `Enhancing gravitational-wave science with machine learning', \textit{Machine Learning: Science and Technology}, vol.\ 2, no.\ 1, 011002, doi:10.1088/2632-2153/abb93a. \url{https://doi.org/10.1088/2632-2153/abb93a}.*
\item Cuoco, E, Cavagli\`{a}, M, Heng, IS, Keitel, D \& Messenger, C 2024, `Applications of machine learning in gravitational-wave research with current interferometric detectors', arXiv:2412.15046. \url{https://arxiv.org/abs/2412.15046}.*
\item David, R, Duke, J, Jain, A, Janapa Reddi, V, Jeffries, N, Li, J, Kreeger, N, Nappier, I, Natraj, M, Regev, S, Rhodes, R, Wang, T \& Warden, P 2021, `TensorFlow Lite Micro: embedded machine learning on TinyML systems', in \textit{Proceedings of Machine Learning and Systems (MLSys)}, vol.\ 3, pp.\ 800--811, arXiv:2010.08678. \url{https://arxiv.org/abs/2010.08678}.*
\item Dax, M, Green, SR, Gair, J, Macke, JH, Buonanno, A \& Sch\"{o}lkopf, B 2021, `Real-time gravitational-wave science with neural posterior estimation', \textit{Physical Review Letters}, vol.\ 127, no.\ 24, 241103, doi:10.1103/PhysRevLett.127.241103. \url{https://doi.org/10.1103/PhysRevLett.127.241103}.*
\item Gabbard, H, Williams, M, Hayes, F \& Messenger, C 2018, `Matching matched filtering with deep networks for gravitational-wave astronomy', \textit{Physical Review Letters}, vol.\ 120, no.\ 14, 141103, doi:10.1103/PhysRevLett.120.141103. \url{https://doi.org/10.1103/PhysRevLett.120.141103}.*
\item Gebhard, TD, Kilbertus, N, Harry, I \& Sch\"{o}lkopf, B 2019, `Convolutional neural networks: a magic bullet for gravitational-wave detection?', \textit{Physical Review D}, vol.\ 100, no.\ 6, 063015, doi:10.1103/PhysRevD.100.063015. \url{https://doi.org/10.1103/PhysRevD.100.063015}.*
\item George, D \& Huerta, EA 2018, `Deep learning for real-time gravitational-wave detection and parameter estimation: results with Advanced LIGO data', \textit{Physics Letters B}, vol.\ 778, pp.\ 64--70, doi:10.1016/j.physletb.2017.12.053. \url{https://doi.org/10.1016/j.physletb.2017.12.053}.*
\item Gholami, A, Kim, S, Dong, Z, Yao, Z, Mahoney, MW \& Keutzer, K 2021, `A survey of quantization methods for efficient neural network inference', arXiv:2103.13630. \url{https://arxiv.org/abs/2103.13630}.*
\item Hinton, G, Vinyals, O \& Dean, J 2015, `Distilling the knowledge in a neural network', arXiv:1503.02531. \url{https://arxiv.org/abs/1503.02531}.*
\item Jacob, B, Kligys, S, Chen, B, Zhu, M, Tang, M, Howard, A, Adam, H \& Kalenichenko, D 2018, `Quantization and training of neural networks for efficient integer-arithmetic-only inference', in \textit{Proceedings of the IEEE/CVF Conference on Computer Vision and Pattern Recognition (CVPR)}, pp.\ 2704--2713, arXiv:1712.05877. \url{https://arxiv.org/abs/1712.05877}.*
\item Krastev, PG 2020, `Real-time detection of gravitational waves from binary neutron-star mergers using artificial neural networks', \textit{Physics Letters B}, vol.\ 803, 135330, doi:10.1016/j.physletb.2020.135330. \url{https://doi.org/10.1016/j.physletb.2020.135330}.*
\item Murphy, D, Ulyanov, A, McBreen, S, Doyle, M, Dunwoody, R, Mangan, J, Erkal, J, Hibbert, B, Marshall, A, Martin-Carrillo, A, Mathews, J, Murphy, D, Norman, C, O'Toole, L, Reilly, J, Salmon, L, Sherwin, J, Shortt, B, Thompson, J, Walsh, S, Wall, S, Wells, J, Wesseling, F \& Hanlon, L 2021, `A compact instrument for gamma-ray burst detection on a CubeSat platform (EIRSAT-1 GMOD)', \textit{Experimental Astronomy}, vol.\ 52, pp.\ 1--32, doi:10.1007/s10686-021-09779-9. \url{https://doi.org/10.1007/s10686-021-09779-9}.*
\item Qi, H, Chen, H-Y, Harms, J, Cusin, G \& Sesana, A 2024, `Improving early detection of gravitational waves from binary neutron stars using CNNs and FPGAs', arXiv:2409.05068. \url{https://arxiv.org/abs/2409.05068}.*
\item Sch\"{a}fer, MB, Zelenka, O, Nitz, AH, Wang, H, Wu, S, Guo, Z-K, Cao, Z, Ren, Z, Nousi, P, Stergioulas, N, Iosif, P, Koloniari, AE, Tefas, A, Passalis, N, Salemi, F, Vedovato, G, Klimenko, S, Mishra, T, Br\"{u}gmann, B, Cuoco, E, Huerta, EA, Messenger, C \& Ohme, F 2023, `MLGWSC-1: the first machine learning gravitational-wave search mock data challenge', \textit{Physical Review D}, vol.\ 107, no.\ 2, 023021, doi:10.1103/PhysRevD.107.023021. \url{https://doi.org/10.1103/PhysRevD.107.023021}.*
\end{enumerate}

\subsection*{Additional references cited but not part of the 18 in-depth included set}

\begin{enumerate}[label=\arabic*., start=19, leftmargin=2em]
\item Lin, J, Chen, W-M, Lin, Y, Cohn, J, Gan, C \& Han, S 2020, `MCUNet: tiny deep learning on IoT devices', in \textit{Advances in Neural Information Processing Systems (NeurIPS)}, vol.\ 33, pp.\ 11711--11722, arXiv:2007.10319. \url{https://arxiv.org/abs/2007.10319}
\item Messick, C, Blackburn, K, Brady, P, Brockill, P, Cannon, K, Cariou, R, Caudill, S, Chamberlin, SJ, Creighton, JDE, Everett, R, Hanna, C, Keppel, D, Magee, RN, Meacher, D, Messenger, C, Mohapatra, S, Nissanke, S, Privitera, S, Sathyaprakash, BS, Singer, L, Tapai, T, Vitale, S, Wade, M, Wade, L, Weinstein, A \& Williamson, AR 2017, `Analysis framework for the prompt discovery of compact binary mergers in gravitational-wave data', \textit{Physical Review D}, vol.\ 95, no.\ 4, 042001, doi:10.1103/PhysRevD.95.042001. \url{https://doi.org/10.1103/PhysRevD.95.042001}
\item Sze, V, Chen, Y-H, Yang, T-J \& Emer, JS 2017, `Efficient processing of deep neural networks: a tutorial and survey', \textit{Proceedings of the IEEE}, vol.\ 105, no.\ 12, pp.\ 2295--2329, doi:10.1109/JPROC.2017.2761740. \url{https://doi.org/10.1109/JPROC.2017.2761740}
\item Warden, P \& Situnayake, D 2019, \textit{TinyML: Machine Learning with TensorFlow Lite on Arduino and Ultra-Low-Power Microcontrollers}, O'Reilly Media, Sebastopol, CA. ISBN 978-1492052043.
\item Wasif, R 2026, \textit{Quantisation-Induced SNR Degradation: Toward Signal-Preserving TinyML for Gravitational Wave Triggers}, Research Proposal, COMP4450, Australian National University.
\end{enumerate}

\end{document}
